---
title: "Circular Data Analysis and Visualizations with R"
subtitle: "Sites of Circadian Clock Neuron Plasticity Mediate Sensory Integration and Entrainment"
abstract: 'The data analysis and visualizations in this report have been developed for the paper *Sites of Circadian Clock Neuron Plasticity Mediate Sensory Integration and Entrainment*, authored by: Fern&aacute;ndez MP^[Correspondence should be addressed to MP Fern&aacute;ndez and OT Shafer.], Pettibone HL, Roell CJ, Davey C, Huynh KV, Lennox S, Kostadinov B and Shafer OT^1^.'
author: ''
date: "2/14/2020"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We use [R](https://www.r-project.org/) and [RStudio](https://rstudio.com/) for all data analysis and visualizations. The phase data are available in the paper cited above, and for the purpose of this data analysis we recorded it in the file `phases.csv`. The source [R Markdown](https://rmarkdown.rstudio.com/) document (with `Rmd` extension) containing all R code is also available in the [GitHub repository](https://github.com/bkostadi/circular_analysis). 

The key R package that we use for all circular data analysis and visualizations is [**circular**](https://www.rdocumentation.org/packages/circular/versions/0.4-93). All R packages must be installed first before they are loaded with `library()`. 

```{r,warning=FALSE,message=FALSE}
# load the R packages
library(readr) # for reading data
library(circular) # main computing package
library(scales) # for plot colors support
# load the phase data, which is available in the cited paper
phases <- read_csv("phases.csv") 
```


```{r}
# take phases modulo 24 to remove negative phases and convert to radians
radphases <- (phases %% 24)*pi/12 # hours in 00:24 and then radians in 0:2pi
cirradphases <- circular(radphases, units = "radians", template = "clock24",
                         modulo="2pi", zero=pi/2) # or zero=2*pi
```


```{r,fig.align='center',fig.width=7,fig.height=7,echo=FALSE,eval=FALSE}
# Test by doing the same conversion using conversion.circular()
mod24phases <- (phases %% 24) # take hours mod 24 to remove negative hours
mod24phases <- circular(mod24phases,units ="hours", modulo = "asis",
                        template = "clock24",zero=24)
test <- conversion.circular(mod24phases,units="radians",template = "clock24",
                            modulo="2pi",zero=pi/2) # or zero=2*pi
# Plotting the data from the two conversions - they should overlap! 
plot(cirradphases[,1], stack = TRUE, col=alpha("blue",0.8), pch = 16, cex=0.8,
     sep = 0.08, shrink = 2, main="")
points(test[,1], stack = TRUE, col=alpha("red",0.8), pch = 16, cex=0.8,
       sep = 0.08, main="")
```


# Plot 1: PdfGal4, uasUNC5 and pdf>UNC5

In Figure \ref{fig:plot1}, we show the raw circular data plots and rose diagrams of **PdfGal4**, **uasUNC5** and **pdf>UNC5**, based on the angles in radians, obtained from the phases. 

```{r plot1,message=FALSE,warning=FALSE,fig.align='center',fig.width=7,fig.height=7,fig.cap="Circular plots and rose diagrams of PdfGal4, uasUNC5 and pdf>UNC5.",fig.pos="h"}
# Plot 1
par(mar=c(rep(1,4))) # set up plot margins
# plot the circular data as a histogram with stack= TRUE
plot(cirradphases[,1], stack = TRUE, pch = 21, bg = alpha("gray77",0.8), col="darkgray", 
     cex=0.8, sep = 0.08, shrink = 2.2,main="PdfGal4, uasUNC5 and pdf>UNC5",
     cex.main=0.8) # plot PdfGal4 [,1]

# calculate the circular means and point to them by arrows
circ.mean1<-mean(cirradphases[,1],na.rm = TRUE)
circ.mean2<-mean(cirradphases[,2],na.rm = TRUE)
circ.mean3<-mean(cirradphases[,3],na.rm = TRUE)

# adding points for uasUNC5 [,2] and pdf>UNC5 [,3]
points(cirradphases[,2],stack = TRUE, pch = 21, bg = alpha("gray60",0.8), 
       col = "darkgray", cex=0.8, sep=0.08) 
points(cirradphases[,3],stack=TRUE, pch = 21, bg = alpha("firebrick",0.8), 
       col = "firebrick", cex=0.8, sep=0.08)

# adding arrows to circular means
arrows.circular(circ.mean1, col = alpha("gray77",0.6), lwd=3) # arrow to circ. mean 
arrows.circular(circ.mean2, col = alpha("gray60",0.6), lwd=3) # arrow to circ. mean
arrows.circular(circ.mean3, col = alpha("firebrick",0.6), lwd=3) # arrow to circ. mean
legend(-0.3,-0.2,legend=c("Pdf-Gal4", "uas-UNC5","Pdf>UNC5"),
       col=c(alpha("gray77",0.9), alpha("gray60",0.8),alpha("firebrick",0.8)),
       pch=16,cex=0.8,box.lty=0)

# Rose diagrams
rose.diag(cirradphases[,1],bins=12,col=alpha("gray77",0.3),
          cex=0.8,prop=0.95,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,2],bins=12,col=alpha("gray60",0.3),
          cex=0.8,prop=0.95,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,3],bins=12,col=alpha("firebrick",0.3),
          cex=0.8,prop=0.95,add=TRUE,ticks=FALSE)
```


# Plot 2: Pdf-Gal4, uas-DBTL and Pdf>DBTL

In Figure \ref{fig:plot2}, we show the raw circular data plots and rose diagrams of **Pdf-Gal4**, **uas-DBTL** and **Pdf>DBTL**, based on the angles in radians, obtained from the phases.

```{r plot2,message=FALSE,warning=FALSE,fig.align='center',fig.width=7,fig.height=7,fig.cap="Circular plots and rose diagrams of Pdf-Gal4, uas-DBTL and Pdf>DBTL.",fig.pos="h"}
# Plot 2
par(mar=c(rep(1,4))) # set plot margins
# plot the circular data for Pdf-Gal4 [1] as a histogram with stack = TRUE
plot(cirradphases[,1], stack = TRUE, pch = 21, bg = alpha("gray77",0.8), col="darkgray", 
     cex=0.8, sep = 0.08, shrink = 2.2, 
     main="Pdf-Gal4, uas-DBTL and Pdf>DBTL",
     cex.main=0.8) # plot PdfGal4 [,1]

# calculate the circular means and point to them by arrows
circ.mean1<-mean(cirradphases[,1],na.rm = TRUE) # Pdf-Gal4 [1]
circ.mean4<-mean(cirradphases[,4],na.rm = TRUE) # uas-DBTL [4]
circ.mean6<-mean(cirradphases[,6],na.rm = TRUE) # Pdf>DBTL [6]

# adding points for uas-DBTL [4] and pdf>DBTL [6]
points(cirradphases[,4],stack = TRUE, pch = 21, bg = alpha("gray45",0.8), 
       col = "darkgray", cex=0.8, sep=0.08) 
points(cirradphases[,6],stack=TRUE, pch = 21, bg = alpha("gold3",0.8), col = "gold3",
       cex=0.8, sep=0.08)

# adding arrows to circular means
arrows.circular(circ.mean1, col = alpha("gray77",0.6), lwd=3) # arrow to circ. mean 
arrows.circular(circ.mean4, col = alpha("gray45",0.6), lwd=3) # arrow to circ. mean
arrows.circular(circ.mean6, col = alpha("gold3",0.6), lwd=3) # arrow to circ. mean
legend(-0.3,-0.2,legend=c("Pdf-Gal4", "uas-DBTL","Pdf>DBTL"),
       col=c(alpha("gray77",0.9), alpha("gray45",0.8),alpha("gold3",0.8)),
       pch=16,cex=0.8,box.lty=0)

# Rose diagrams
rose.diag(cirradphases[,1],bins=12,col=alpha("gray77",0.3),
          cex=0.8,prop=0.88,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,4],bins=12,col=alpha("gray45",0.3),
          cex=0.8,prop=0.88,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,6],bins=12,col=alpha("gold3",0.3),
          cex=0.8,prop=0.88,add=TRUE,ticks=FALSE)
```


# Plot 3: Pdf>DBTL, uas-UNC5 and pdf>DBTL,UNC5

In Figure \ref{fig:plot3}, we show the raw circular data plots and rose diagrams of **Pdf>DBTL**, **uas-UNC5** and **pdf>DBTL,UNC5** based on the angles in radians, obtained from the phases.

```{r plot3,message=FALSE,warning=FALSE,fig.align='center',fig.width=7,fig.height=7,fig.cap="Circular plots and rose diagrams of Pdf>DBTL, uas-UNC5 and pdf>DBTL,UNC5.",fig.pos="h"}
# Plot 3
par(mar=c(rep(1,4)))
# plot the circular data for Pdf>DBTL [6] as a histogram with stack = TRUE
plot(cirradphases[,6], stack = TRUE, pch = 21, bg = alpha("gold3",0.8), col="gold3", 
     cex=0.8, sep = 0.08, shrink = 2.2, 
     main="Pdf>DBTL, uas-UNC5 and pdf>DBTL,UNC5",
     cex.main=0.8)

# calculate the circular means and point to them by arrows
circ.mean6<-mean(cirradphases[,6],na.rm = TRUE) # Pdf>DBTL [6]
circ.mean2<-mean(cirradphases[,2],na.rm = TRUE) # uas-UNC5 [2]
circ.mean7<-mean(cirradphases[,7],na.rm = TRUE) # pdf>DBTL,UNC5 [7]

# adding points for uas-UNC5 [2] and pdf>DBTL,UNC5 [7]
points(cirradphases[,2],stack = TRUE, pch = 21, bg = alpha("gray60",0.8), 
       col = "darkgray", cex=0.8, sep=0.08) 
points(cirradphases[,7],stack=TRUE, pch = 21, bg = alpha("gold4",0.8), col = "gold4",
       cex=0.8, sep=0.08)

# adding arrows to circular means
arrows.circular(circ.mean6, col = alpha("gold3",0.6), lwd=3) # arrow to circ. mean 
arrows.circular(circ.mean2, col = alpha("gray60",0.6), lwd=3) # arrow to circ. mean
arrows.circular(circ.mean7, col = alpha("gold4",0.6), lwd=3) # arrow to circ. mean
legend(-0.4,-0.2,legend=c("Pdf>DBTL", "uas-UNC5","pdf>DBTL,UNC5"),
       col=c(alpha("gold3",0.8), alpha("gray60",0.8),alpha("gold4",0.8)),
       pch=16,cex=0.8,box.lty=0)

# Rose diagrams
rose.diag(cirradphases[,6],bins=12,col=alpha("gold3",0.3),
          cex=0.8,prop=0.95,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,2],bins=12,col=alpha("gray60",0.3),
          cex=0.8,prop=0.95,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,7],bins=12,col=alpha("gold4",0.3),
          cex=0.8,prop=0.95,add=TRUE,ticks=FALSE)
```


# Plot 4: Pdf>DBTL, uas-DBTL and pdf>DBTL,UNC5

In Figure \ref{fig:plot4}, we show the raw circular data plots and rose diagrams of Pdf>DBTL, uas-DBTL and pdf>DBTL,UNC5, based on the angles in radians, obtained from the phases.

```{r plot4,message=FALSE,warning=FALSE,fig.align='center',fig.width=7,fig.height=7,fig.cap="Circular plots and rose diagrams of Pdf>DBTL, uas-DBTL and pdf>DBTL,UNC5.",fig.pos="h"}
# Plot 4
par(mar=c(rep(1,4))) # set margins for the plots
# plot the circular data for Pdf>DBTL [6] as a histogram with stack = TRUE
plot(cirradphases[,6], stack = TRUE, pch = 21, bg = alpha("gold3",0.8), col="gold3", 
     cex=0.8, sep = 0.08, shrink = 2.2, 
     main="Pdf>DBTL, uas-DBTL and pdf>DBTL,UNC5",
     cex.main=0.8)

# calculate the circular means and point to them by arrows
circ.mean6<-mean(cirradphases[,6],na.rm = TRUE) # Pdf>DBTL [6]
circ.mean4<-mean(cirradphases[,4],na.rm = TRUE) # uas-DBTL [4]
circ.mean7<-mean(cirradphases[,7],na.rm = TRUE) # pdf>DBTL,UNC5 [7]

# adding points for uas-DBTL [4] and pdf>DBTL,UNC5 [7]
points(cirradphases[,4],stack = TRUE, pch = 21, bg = alpha("gray45",0.8), 
       col = "darkgray", cex=0.8, sep=0.08) 
points(cirradphases[,7],stack=TRUE, pch = 21, bg = alpha("gold4",0.8), col = "gold4",
       cex=0.8, sep=0.08)

# adding arrows to circular means
arrows.circular(circ.mean6, col = alpha("gold3",0.6), lwd=3) # arrow to circ. mean 
arrows.circular(circ.mean4, col = alpha("gray45",0.6), lwd=3) # arrow to circ. mean
arrows.circular(circ.mean7, col = alpha("gold4",0.6), lwd=3) # arrow to circ. mean
legend(-0.4,-0.2,legend=c("Pdf>DBTL", "uas-DBTL","pdf>DBTL,UNC5"),
       col=c(alpha("gold3",0.8), alpha("gray45",0.8),alpha("gold4",0.8)),
       pch=16,cex=0.8,box.lty=0)

# Rose diagrams
rose.diag(cirradphases[,6],bins=12,col=alpha("gold3",0.3),
          cex=0.8,prop=0.88,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,4],bins=12,col=alpha("gray45",0.3),
          cex=0.8,prop=0.88,add=TRUE,ticks=FALSE)
rose.diag(cirradphases[,7],bins=12,col=alpha("gold4",0.3),
          cex=0.8,prop=0.88,add=TRUE,ticks=FALSE)
```





